Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-07-03 pituba/article/article_spe.tex
improvs

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}
Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

>>>>

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}

Assuming reservoir temperature to be sufficiently above gas critical temperature for negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}
